Automatic Matching of Bullet Lands

### 2016-10-26 imaging-paper.Rnw
More updates

Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of convictions in the United States criminal justice system. Firearms identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of firearms identification evidence~\citep{giannelli:2011}.

@@

The purpose of our paper is to present an automatic matching routine that allows for a completely objective assessment of the quality of a match. Our algorithm is fully open source and available on GitHub \citep{x3prplus}. This transparency allows for a greater understanding of the individual steps involved in the bullet matching process, and allows other forensic examiners, as well as outside observers, examine the factors that separate known bullet matches from non-matches. We have chosen to perform the matching on a land-to-land level, rather than bullet-to-bullet. Although doing so introduces an implicit assumption of independence between lands, assuming independence only serves to make the task more challenging.
The original paper on the complete Hamby study already reports the successful use of several computer-assisted methods; aside from a zero false positive rate, no further details on a false-negative error rate for bullets are given nor are error rates for land-to-land matches mentioned. 
% Hamby study: automated systems were successfully used for matching:
%  Intelligent Automation's SciClops - Dr. Ben Bachrach (Maryland, United States);
% Automated Land Identification System (ALIS) - Mr. Tsuneo Uchiyama (Tokyo, Japan);
% Integrated Ballistics Identification System (IBIS) – Mr. Robert Thompson (California, United States);
% BulletTRAX-3DTM - Forensic Technology Scientists (Montreal, Canada) (2 sets)

A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. This study is based on another set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available. Modern automated techniques using 3D images have also been conducted by \citet{riva:2014}, but this work applies to striae found on cartridge cases and not bullets. Other work on 3D images has been conducted by \citet{petraco:2012}, which also describes work carried out on cartridge cases, as well as screwdriver striation patterns, with promising results. We aim to use similar ideas of automated 3D image analysis, and apply this work to bullet surfaces with the modern tools enabled by the R statistical software.

>>>>

Firearm examination is a forensic tool used to help the court determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of convictions in the United States criminal justice system. Firearms identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of firearms identification evidence~\citep{giannelli:2011}.

@@

The purpose of our paper is to present an automatic matching routine that allows for a completely objective assessment of the quality of a match. While we assess the performance of the algorithm in terms of a binary decision of match vs. non-match using a 50\% probability cut-off, our primary goal is to highlight the features that are statistically associated with matches and non-matches, and to provide a quantitative assessment of this association. In a real-world application of our algorithm, the raw scores would need further analysis and scrutiny, and it is likely a 50\% cut-off would be an inappropriate choice on the basis of reasonable doubt.

Our algorithm is fully open source and available on GitHub \citep{x3prplus}. This transparency allows for a greater understanding of the individual steps involved in the bullet matching process, and allows other forensic examiners, as well as outside observers, examine the factors that separate known bullet matches from non-matches. We have chosen to perform the matching on a land-to-land level, rather than bullet-to-bullet. Although doing so introduces an implicit assumption of independence between lands, assuming independence only serves to make the task more challenging.
The original paper on the complete Hamby study already reports the successful use of several computer-assisted methods; aside from a zero false positive rate, no further details on a false-negative error rate for bullets are given nor are error rates for land-to-land matches mentioned. 
% Hamby study: automated systems were successfully used for matching:
%  Intelligent Automation's SciClops - Dr. Ben Bachrach (Maryland, United States);
% Automated Land Identification System (ALIS) - Mr. Tsuneo Uchiyama (Tokyo, Japan);
% Integrated Ballistics Identification System (IBIS) – Mr. Robert Thompson (California, United States);
% BulletTRAX-3DTM - Forensic Technology Scientists (Montreal, Canada) (2 sets)

Work has been done to quantify toolmark similarity by \citet{lock2013significance}. This algorithm determines an optimal matching window between two toolmark signatures, and then performs a set of both coordinated and independent shifts. Given a match, the coordinated shifts would be expected to yield correlation values higher than those obtained from independent shifts. This is assessed using a Mann-Whitney U Statistic. A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. This study is based on another set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available.

Modern automated techniques using 3D images have also been conducted by \citet{riva:2014}, but this work applies to striae found on cartridge cases and not bullets. This algorithm performs alignment of striae by rotation of the XY plane, which is not generalizable to bullets in which the XY plane is not flat. Other work on 3D images has been conducted by \citet{petraco:2012}, which also describes work carried out on cartridge cases, as well as screwdriver striation patterns, with promising results. We aim to use similar ideas of automated 3D image analysis, and apply this work to bullet surfaces with the modern tools enabled by the R statistical software.
